Interactive Visualization of Hierarchically Structured Data

### 2016-12-04 treelapse.Rmd
Revisions to problem background

- DOI Trees
     + definition and (microbiome) use case: for a single sample, where are
       most bacteria?
     + principles it relies on: just DOI tree.
     + This is the only visualization that is completely thought through from
       before (but which as far as we can tell hasn't had easily accessible
       implementations).
     + well not exactly...we are also encoding abundance in edge widths / node
       sizes
- DOI Sankeys
     + Definition and use case: treatment vs. control in tree structured data,
       but using DOI principle
     + Exploratory analog of structSSI
     + Minor note: we use the colorbrewer palette for simpler readability
- Timebox Trees
     + Combines a few more ideas: linking and focus + context
     + definition and explanation of how these ideas being implemented
              * timebox recap: we're able to highlight lines passing through
                intersection of boxes (this is focus and context)
              * Link the display between tree and time series. Interpretation:
                Position on tree conditional on time series
                shape. Alternatively, database query on one field, returning
                over all of them.
              * Focus and context / pan and zoom using visual scent [so we have
                two focus and context ideas, and one linking idea, in this one
                display]
      + Primary use case this is designed for -- we notice a pattern in how a
        clump of microbial series look like, and we try to figure out which
        subtrees have this behavior (or, if that behavior is appear at
        different places on the tree).
- Treeboxes
       + Describe as the natural analog of timebox trees
       + Intentionally avoided queries on both tree and time series
       + Example use case: we can see how different parts of the tree have
         different series.
               * Or, we can look at all the phyla / orders at once
               * Or, we can observe if there is any competition within a
                 particular clade
       + Conditional on coming from a part of a tree, what are the observed
         series?

@@

In this work, we describe methods for visualization of data structured
along trees, especially hierarchically structured collections of time
series. Influenced by Degree-of-Interest trees [@heer2004doitrees],
linked scatterplot brushing [@becker1987brushing], and timeboxes
[@hochheiser2004dynamic], we have adapted the fundamental
visualization principles of "focus + context" and "linking" to study
questions that arise when analyzing tree structured data. Our
motivating application is to the analysis of microbial time series,
where an evolutionary tree relating microbes is available a
priori. However, we have identified common problem types where, if a
tree is not directly available, it can be constructed from data and
then studied using our techniques.

@@

An abstraction that unifies thinking through both the tree structured
differential abundance and microbial dynamic problems is to identify the data
with a collection of random variables indexed by nodes in a prespecified tree
structure. In the differential abundance problem, each random variable itself
lives in $\mathbb{R}^{G}$ where $G$ is the number of groups being
compared. Each coordinate represents the average value across one of the
groups, and the node exhibits differential abundance when the coordinates are
not drawn from the same distribution. In the microbial dynamics problem, each
random variable is instead a time series, living in $\mathbb{R}^{T}$.

In both these applications, we constrain the values of parent nodes according
to the value of the children nodes: we define the value at each node to be
either the sum or average of all descendant tips, but it is possible to imagine
situations where the internal nodes are drawn from their own distribution,
unconstrained by descendants. In general, analysis in this abstraction focuses
on describing the distribution of these random variables as a function of their
position in specified subsets of the tree. The essential difficulty in these
problems is high-dimensionality -- there are many tree nodes, each holding a
vector-valued random variable. Even simply navigating across the tree and
comparing coordinates in the observed variables is a challenge; ideally we
could construct a succinct representation of the essential covariation across
subtrees and coordinates.

This framework suggests other potential application areas, not all of which
have a priori known tree structures. For example, collections of spatially
indexed time series are frequently encountered in practice -- consider energy
consumption, product sales, or high school dropout rates across regional
districts. This type of data has an implicit tree structure -- at the top level
are different states, while at the bottom are individual census tracts,
say. Analysis here revolves around the question of how variation across series
is related to their geographic position.  An example using this type of data is
given in the Zillow home prices case study in Section (todo: insert section
number).

Alternatively, if this type of hierarchical contextual information is not
available, a tree structure can be learned directly from the data. This can be
most directly achieved by learning a hierarchical clustering on the original
series. If there is contextual information, but it is not hierarchical, it is
possible to setup a supervised problem that uses context to predict features of
the time series. We can retrieve a tree by using a regression tree directly
(cite: rpart) or extracting a regression tree from a more complex supervised
model (cite: regression tree extraction methods). Analysis then focuses on how
different partitions of the covariate (contextual) space relate to observed
time series. This approach is described in the UCI bikesharing demand case
study below.

Finally, note that, while we have focused on time series valued nodes, all of
this discussion could be translated to studying high dimensional data via
parallel coordinates (todo: cite parallel coordinates). The usual parallel
coordinates challenges remain, mainly selecting scales for and ordering the
different coordinates, but the linking and focus + context can still be
employed this setting.

@@

Now that we have specified the essential questions of interest, we survey some
ideas from the visualization literature that can be applied to answer
them. Again, the core difficulty is one of high dimensionality, so it should be
no surprise that the techniques we adapt come from the literature on
high-dimensional data visualization, namely, work on the focus + context and
linking.

The focus + context principle is that it is possible to study large collections
of visual elements at multiple scales, and these scales can be juxtaposed by
"focusing" on a few elements of interest while maintaining the "context" of the
coarser-scale background. A simple example of this idea is to include a search
box that highlights samples corresponding to matching terms (the focus) while
muting those that don't (the context). Two more sophisticated methods anchored
in this idea are timeboxes and Degree-of-Interest (DOI) trees; both are central
to the proposals in treelapse \@cite(hochheiser2004dynamic,
heer2004doitrees). In timeboxes, a collection of time series are graphically
queried using interactive brushes. Series that pass through all of the
user-specified brushes are highlighted, and the rest are faded into the
background. Hence, time series meeting the constraints imposed by the brushes
are focused, while the remainder remain present, though de-emphasized, as
context. This method can be interpreted programmatically as the visual version
of a database query, or probabilistically as the conditional distribution for
the full series, given it passes through certain bounds.

@@

### Dynamics in Housing Prices ###

@@

### Sources of Variation in Bikesharing Demand ###

@@

All these potential extensions aside, we have built an easily
accessible R package for visualization techniques in a very specific
methodology problem – analysis of differential abundance and dynamics
in hierarchical structured data – that appears in a variety of
application domains. We have leveraged (Vaidyanathan et al.  2014) to
create exploratory visualizations while still deep in data analysis;
in this way our work is a departure from the culture of polished,
journalistic visualizations prioritized by the D3 community. Finally,
we have given a series of examples to demonstrate how the general
visualization techniques of focus + context and linked brushing can be
practically incorporated into a range of practical analysis workflows,
from studying the impact of microbes on human health to better
allocating units in commuter bikesharing systems.

>>>>

- DOI Trees
     + definition and (microbiome) use case: for a single sample, where are
       most bacteria?
     + principles it relies on: just DOI tree.
     + This is the only visualization that is completely thought through from
       before (but which as far as we can tell hasn't had easily accessible
       implementations).
     + well not exactly...we are also encoding abundance in edge widths / node
       sizes
- DOI Sankeys
     + Definition and use case: treatment vs. control in tree structured data,
       but using DOI principle
     + Exploratory analog of structSSI
     + Minor note: we use the colorbrewer palette for simpler readability
- Timebox Trees
     + Combines a few more ideas: linking and focus-plus-context
     + definition and explanation of how these ideas being implemented
              * timebox recap: we're able to highlight lines passing through
                intersection of boxes (this is focus and context)
              * Link the display between tree and time series. Interpretation:
                Position on tree conditional on time series
                shape. Alternatively, database query on one field, returning
                over all of them.
              * Focus and context / pan and zoom using visual scent [so we have
                two focus and context ideas, and one linking idea, in this one
                display]
      + Primary use case this is designed for -- we notice a pattern in how a
        clump of microbial series look like, and we try to figure out which
        subtrees have this behavior (or, if that behavior is appear at
        different places on the tree).
- Treeboxes
       + Describe as the natural analog of timebox trees
       + Intentionally avoided queries on both tree and time series
       + Example use case: we can see how different parts of the tree have
         different series.
               * Or, we can look at all the phyla / orders at once
               * Or, we can observe if there is any competition within a
                 particular clade
       + Conditional on coming from a part of a tree, what are the observed
         series?

@@

In this work, we describe methods for visualization of data structured
along trees, especially hierarchically structured collections of time
series. Influenced by Degree-of-Interest trees [@heer2004doitrees],
linked scatterplot brushing [@becker1987brushing], and timeboxes
[@hochheiser2004dynamic], we have adapted the fundamental
visualization principles of "focus-plus-context" and "linking" to study
questions that arise when analyzing tree structured data. Our
motivating application is to the analysis of microbial time series,
where an evolutionary tree relating microbes is available a
priori. However, we have identified common problem types where, if a
tree is not directly available, it can be constructed from data and
then studied using our techniques.

@@

To unify the tree-structured differential abundance and microbial
dynamic problems, we identify the data with a collection of random
variables indexed by nodes in a prespecified tree structure. In the
differential abundance problem, each random variable lives in
$\mathbb{R}^{G}$ where $G$ is the number of groups being
compared. Each coordinate represents the abundance for that group, and
a node exhibits differential abundance when the coordinates are drawn
from different distributions. On the other hand, in the microbial
dynamics problem, each random variable is a time series, living in
$\mathbb{R}^{T}$.

In both of these applications, we constrain the values of parent nodes
according to the value of the children nodes: we define the value at
each node to be either the sum or average of all descendant
tips. However, it is possible to imagine situations where the internal
nodes are drawn from their own distribution, unconstrained by
descendants. In general, analysis in this abstraction focuses on
describing the distribution of these random variables as a function of
their position across subsets of the tree. The essential difficulty in
these problems is high-dimensionality -- there are many tree nodes,
each holding a vector-valued random variable. Even simply navigating
across the tree and comparing coordinates in the observed variables is
a challenge; ideally we could construct a succinct representation of
the essential covariation across subtrees and coordinates.

This framework suggests other potential application areas, not all of
which have a priori known tree structures. For example, collections of
spatially-indexed time series are frequently encountered in practice
-- consider energy consumption, product sales, or high school dropout
rates across regional districts. This type of data has an implicit
tree structure -- at the top level are different states, while at the
bottom are individual census tracts, say. Analysis here revolves
around the question of how variation across time series is related to
their geographic position.  A case study to this type of data is given
in Section \@ref(zillow-study).

Alternatively, if this type of hierarchical contextual information is
not directly available, a tree structure can be learned from the
data. This could be achieved by learning a hierarchical clustering on
the original series. Further, if contextual information is available,
but it is not hierarchical, it is possible to setup a supervised
problem that uses context to predict features of the time series. We
can construct a tree by applying a tree-based classifier
[@breiman1984classification] or extracting a regression tree from a
more complex supervised model [@boz2002extracting,
@saito2002extracting]. Analysis then focuses on how different
partitions of the contextual, covariate space relate to observed time
series. This approach is described in Section \@ref(bikesharing-study).

Finally, note that, while we have focused on time series valued nodes,
all of this discussion could be translated to studying
high-dimensional data via parallel coordinates
[@inselberg1991parallel]. The usual parallel coordinates challenges
remain, mainly selecting scales for and ordering across the different
coordinates, but the linking and focus-plus-context can still be
employed this setting.

@@

Now that we have specified the essential questions of interest, we survey some
ideas from the visualization literature that can be applied to answer
them. Again, the core difficulty is one of high dimensionality, so it should be
no surprise that the techniques we adapt come from the literature on
high-dimensional data visualization, namely, work on the focus-plus-context and
linking.

The focus-plus-context principle is that it is possible to study large collections
of visual elements at multiple scales, and these scales can be juxtaposed by
"focusing" on a few elements of interest while maintaining the "context" of the
coarser-scale background. A simple example of this idea is to include a search
box that highlights samples corresponding to matching terms (the focus) while
muting those that don't (the context). Two more sophisticated methods anchored
in this idea are timeboxes and Degree-of-Interest (DOI) trees; both are central
to the proposals in treelapse \@cite(hochheiser2004dynamic,
heer2004doitrees). In timeboxes, a collection of time series are graphically
queried using interactive brushes. Series that pass through all of the
user-specified brushes are highlighted, and the rest are faded into the
background. Hence, time series meeting the constraints imposed by the brushes
are focused, while the remainder remain present, though de-emphasized, as
context. This method can be interpreted programmatically as the visual version
of a database query, or probabilistically as the conditional distribution for
the full series, given it passes through certain bounds.

@@

### Dynamics in Housing Prices {#zillow-study}

@@

### Sources of Variation in Bikesharing Demand {#bikesharing-study}

@@

All these potential extensions aside, we have built an easily
accessible R package for visualization techniques in a very specific
methodology problem – analysis of differential abundance and dynamics
in hierarchical structured data – that appears in a variety of
application domains. We have leveraged (Vaidyanathan et al.  2014) to
create exploratory visualizations while still deep in data analysis;
in this way our work is a departure from the culture of polished,
journalistic visualizations prioritized by the D3 community. Finally,
we have given a series of examples to demonstrate how the general
visualization techniques of focus-plus-context and linked brushing can be
practically incorporated into a range of practical analysis workflows,
from studying the impact of microbes on human health to better
allocating units in commuter bikesharing systems.
